Peptide retention time prediction

### 2014-11-09 retention.tex
Added paragraph and image about elude-ptm

\documentclass[a4paper]{article}
\usepackage{graphicx}
\usepackage{wrapfig}

@@

Although available in a variety of flavours, most of the widely spread MS-based workflows consists of a few common steps. First, the proteins in the mixture of interest are cleaved into peptides using a sequence-specific enzyme such as trypsin. This step is necessary because the mass spectrometers are more sensitive for peptides, and because spectra produced by full-length proteins are difficult to interpret. In the next step, the resulting peptides are separated according to their hydrophobicity on a liquid chromatography (LC) column. As they gradually elute from the column, the peptides are ionized via electrospray ionization and subjected to mass spectrometry analysis. The result of these measurements consists of a set of spectra, which are subsequently used to infer valuable information about the proteins present in the initial sample.

Today a majority of the mass spectrometry (MS) protocols in use are high-throughput, generating extensive amounts of data. This has brought up a whole new set of analytical challenges, making the development of efficient algorithms to manipulate and interpret the data a major area of proteomics research. A particular difficult challenge in this context is posed by the tremendous complexity of the data. Proteins
display a wide range of physico-chemical properties and are often found at very different amounts in complex samples, spanning as much as 12 orders of magnitude in relative
abundance~\cite{Angel2012}. Furthermore, a large proportion of
proteins are subjected to posttranslational modifications. As an
example, a few years ago more than 25,000 phosphorylation sites on
human proteins were already reported~\cite{Lemeer2009}. To manage this formidable diversity,
 increasing efforts are made to extract useful information not only from the spectra, but also from other types of information recorded during MS experiments.

An example of such information is retention time, i.e. the time point when a peptide elutes from the liquid chromatography column. Although the main purpose of liquid chromatography is to reduce the number of peptides
simultaneously injected into the mass spectrometer, and by this
increase the proteome coverage achieved in an experiment. However, besides this, the separation process in itself provides
useful information about the analytes. On the one hand, the instrument
always records the time when a peptide elutes from the chromatography
column, i.e. the retention time. On the other hand, the retention time
can also be estimated by examining the sequence of the peptide. The
availability of both predicted and observed retention times has proven
to be a powerful feature, with several applications in mass
spectrometry-based proteomics \cite{Palmblad2013}.

@@

\begin{wrapfigure}{r}{.45\textwidth}
\vspace{-10pt}
%\begin{figure}[htbp]
\centering
\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
\caption{\label{fig:repr} {\bf Reproducibility of peptide retention time.} Two identical shotgun proteomics experiments were carried out. The retention times of the peptides confidently identified in both runs are displayed.}
%\end{figure}
\vspace{-12pt}
\end{wrapfigure}

@@

@@

@@

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer et al. proposed a predictor based on support vector machines
(SVMs) \cite{klammer2007improving}. Their method required fewer
training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction
accuracies. Subsequently, a more elaborate SVM-based method using a
customized kernel function was proposed \cite{rtpredictImproved}. This
approach provided reasonable prediction accuracy, while requiring as
little as 50 peptides for training.

@@

@@

>>>>

\documentclass[a4paper]{article}
\usepackage[margin=0.5in]{geometry} 
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{multirow}

@@

Although available in a variety of flavours, most of the widely spread MS-based workflows consists of the same main steps. First, the proteins in the mixture of interest are cleaved into peptides using a sequence-specific enzyme such as trypsin. This step is necessary because the mass spectrometers are more sensitive for peptides, and because spectra produced by full-length proteins are difficult to interpret. In the next step, the resulting peptides are separated according to their hydrophobicity on a liquid chromatography (LC) column. As they gradually elute from the column, the peptides are ionized via electrospray ionization and subjected to mass spectrometry analysis. The result of these measurements consists of a set of spectra, which are subsequently used to infer valuable information about the proteins present in the initial sample.

Today a majority of the mass spectrometry (MS) protocols in use are high-throughput, generating extensive amounts of data. This has brought up a whole new set of analytical challenges, making the development of efficient algorithms to interpret this data a major area of proteomics research. A particular difficult challenge in this context is posed by the tremendous complexity of the data. Proteins
display a wide range of physico-chemical properties and are often found at very different amounts in complex samples, spanning as much as 12 orders of magnitude in relative
abundance~\cite{Angel2012}. Furthermore, a large proportion of
proteins are subjected to posttranslational modifications. As an
example, a few years ago more than 25,000 phosphorylation sites on
human proteins were already reported~\cite{Lemeer2009}. To manage this formidable diversity,
 increasing efforts are made to extract useful information not only from the spectra, but also from other types of information recorded during MS experiments. An example of such information is retention time, i.e. the time point when a peptide elutes from the liquid chromatography column.

The main purpose of liquid chromatography is to reduce the number of peptides
simultaneously injected into the mass spectrometer, and by this
increase the proteome coverage achieved in an experiment. However, besides this, the separation process in itself provides
useful information about the analytes. On the one hand, the instrument
always records the time when a peptide elutes from the chromatography
column, i.e. the retention time. On the other hand, the retention time
can also be estimated by examining the sequence of the peptide. The
availability of both predicted and observed retention times has proven
to be a powerful feature, with several applications in mass
spectrometry-based proteomics \cite{Palmblad2013}.

@@

@@

\begin{wrapfigure}{r}{.45\textwidth}
\vspace{-10pt}
%\begin{figure}[htbp]
\centering
\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
\caption{\label{fig:repr} {\bf Reproducibility of peptide retention time.} Two identical shotgun proteomics experiments were carried out. The retention times of the peptides confidently identified in both runs are displayed.}
%\end{figure}
\vspace{-25pt}
\end{wrapfigure}

@@

\begin{table}
 \caption{{\bf Retention time prediction methods.} We
   give a few of the most commonly used retention time prediction methods available.}
 \vspace{.2cm}
 \label{tab:rtmethods}
 \begin{tabular}{lll}

 Method name & Link & References  \\
 \hline
SSRCalc & \url{http://hs2.proteome.ca/SSRCalc/SSRCalcX.html} & \cite{Krokhin2004, Krokhin2006}, \cite{Spicer2007} \\
BioLCCC & \url{http://theorchromo.ru/} & \cite{gorshkov2006} \\
ANN & Link? & \cite{petritis2003 ,petritis2006improved} \\
RTModel, RTPredict &  \url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTPredict.html}, 
 & \cite{rtpredict, rtpredictImproved} \\
&\url{http://ftp.mi.fu-berlin.de/OpenMS/documentation/html/TOPP_RTModel.html}  & \\

Elude & \url{http://elude.sbc.su.se/} & \cite{elude1, elude2} \\
\hline
 \end{tabular} \\
\end{table}

@@

A drawback of the method was however the large amount of training
peptides required, which made the algorithm difficult to retrain for
other chromatographic conditions. This was a serious limitation in the
context of RPLC separation, where a broad variety of chromatographic
setups is commonly used across research laboratories. To address this,
Klammer et al. proposed a predictor based on support vector machines
(SVMs) \cite{klammer2007improving}. Their method required fewer
training peptides and could thus be easily adapted to new
chromatographic conditions, but yielded lower prediction
accuracies. Subsequently, a more elaborate SVM-based method using a
customized kernel function was proposed \cite{rtpredict, rtpredictImproved}. This
approach provided reasonable prediction accuracy, while requiring as
little as 50 peptides for training.

@@

In our group we developed a software package for retention time prediction which we named {\sc
Elude} \cite{elude1, elude2}. One of the main novelty
points of {\sc Elude} wass the calculation of peptide features based on
both a general hydrophobicity index, as well as a retention index
trained for each dataset at hand. This ensured that the algorithm
could be applied to different chromatography setups with minimal loss
in performance. For cases when little training data was available,
{\sc ELUDE} implemented an additional workflow based on a library of
pre-trained retention models. This workflow required only a few tens
of training peptides to select and calibrate the most suitable model
from the library. Both the selection and calibration were performed
using statistical methods robust to outliers.

@@

\begin{figure}[!h]
\centering 
\includegraphics[width=0.9\textwidth]{img/elude-ptm.pdf}
\caption{\label{fig:elude} {\bf {\sc Elude}'s workflow.}  {\sc Elude} encompasses two learning steps. In the first one the training peptides are used to build an in-house retention index. This index, together with additional information, is then employed to calculate a set of 60 numerical features for each peptide. These features are optimally combined during a second training procedure. The resulting retention model can be subsequently used to predict the retention time of other peptides of interest.}
\end{figure}

To address this, in our group we extended the first version of our algorithm {\sc ELUDE} to encompass modified peptides as well. This was straightforward to achieve once we observed that the in-house trained  retention index contained nearly all the information incorporated in the more general hydrophobicity scale. This meant that we could remove all the features based on this index with minimal loss in accuracy. Further, we redefined all features describing a peptide in terms of the in-house retention index alone. Since this index was trained for each dataset at hand, the algorithm became fully trainable for any modified amino acid (Figure \ref{fig:elude}). We evaluated the new version of {\sc ELUDE} on peptides containing artifactual modifications introduced as part of a MS protocol called COmbined
FRActional DIagonal Chromatography (COFRADIC) \cite{Gevaert2002}. Our results showed that  {\sc ELUDE} yielded equally accurate predictions for modified and
unmodified peptides  \cite{elude2}.
